\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-09-06 smurf_paper.tex
add Remo as author, went over section 3 on iterative model calc

\author[Edward~L.~Chapin~et~al.]{
  \parbox[t]{\textwidth}{
    Edward~L.~Chapin$^{1,2}$\thanks{E-mail:~echapin@sciops.esa.int,
      Present address: XMM SOC, ESAC, Apartado 78, 28691 Villanueva de
      la Ca\~nada, Madrid, Spain},
    David~S.~Berry$^{2}$,
    Andrew~G.~Gibb$^{1}$,
    Tim~Jenness$^{2}$,
    Douglas~Scott$^{1}$,
    Frossie~Economou$^2$\thanks{Present address: National Optical
      Astronomy Observatory, 950 N.\ Cherry Avenue, Tucson, AZ 85719, USA}
  }
  \\
  \\
  $^{1}$Dept. of Physics \& Astronomy, University of British Columbia,
  6224 Agricultural Road, Vancouver, B.C. V6T 1Z1, Canada\\
  $^{2}$JointAstronomy Centre, 660 N. A`oh\={o}k\={u} Place, University
  Park, Hilo, Hawaii 96720, USA}

\begin{document}

@@

\begin{abstract}
  The Submillimetre Common User Bolometer Array 2 (\scuba) is an
  instrument operating on the 15-m James Clerk Maxwell Telescope,
  consisting of about 5000 bolometers in each of two simultaneous
  imaging bands centered over 450 and 850\,\micron. The camera is
  scanned across the sky and records data at a rate of 200\,Hz. This
  data volume is much greater than that of previously existing
  submillimetre cameras, and represents a significant data anlaysis
  challenge. We describe the production of SCUBA-2 maps using the
  Sub-Millimetre User Reduction Facility (SMURF) in which we have
  adopted a fast, iterative approach to map-making that enables data
  reduction on single, modern, high-end desktop computers with
  execution times that are shorter than the observing times.  SMURF is
  used in an automated setting both at the telescope for real-time
  feedback to observers, as well as for the production of generic
  science products at the Canadian Astronomical Data Centre. A suite
  of tools are also provided for interactive data analysis.  Three
  detailed case studies are used to: (i) explore convergence
  properties of the map-maker using simple prior constraints (Uranus
  -- a point source); (ii) achieve the white-noise limit of the
  instrument for faint point-source studies (blank-field survey of the
  Lockman Hole); and (iii) demonstrate that our strategy is capable of
  recovering angular scales up to the size of the array footprint
  (approximately 5\,arcmin) for bright extended sources (star-forming
  region M17).
\end{abstract}

@@

The best existing map-making strategies for recovering information on
all angular scales are \emph{maximum likelihood} techniques, in which
the time-series data are expressed as a sampling of the ``true'' map
of the sky plus noise, and then the equation is inverted to estimate
the map as some weighted linear combination of the data that minimizes
the variance.  The first good description of this method appears in
\citet{janssen1992} for the production of maps from the COsmic
Background Explorer (COBE -- the description is relevant despite the
fact that it used a differential radiometer instead of
bolometers). Other descriptions in the CMB literature include
\citet{tegmark1997} and \citet{stompor2002}, and an application to
data from SCUBA is described in \citet{borys2004}.  The downside to
such methods is that they can be both computationally expensive and
have large memory requirements. While for some experimental designs
fast iterative methods for the inversion do exist with minimal memory
usage \citep[e.g.][]{wright1996}, for the more general map-making
problem involving many detectors, the problem is significantly
complicated both by the need to measure the cross power-spectra for
all unique pairs of bolometers, and then to perform the inversion
itself. The most promising maximum-likelihood method that may one day
be applied to SCUBA-2 data is ``SANEPIC'', which successfully reduced
maps from the Balloon-borne Large-Aperture Submillimeter Telescope,
involving hundreds of bolometers, and correctly incorporating
inter-bolometer noise correlations \citep[][]{patanchon2008}.

@@

This paper is divided in the following way. We first describe the
properties of \scuba\ data and scan patterns in Section~\ref{sec:data}
to motivate the data model that we have adopted. Note that most of the
analysis described here applies equally to all of the data taken
to-date with \scuba, including during the Shared-Risks Observing
period (S2SRO) from 2010 February 11 to 2010 March 25 (during which
only single subarrays were available at each of the two observing
wavelengths). A comparison between the S2SRO and fully-commissioned
instrument is given in Section~\ref{sec:s2sro}. Next, the details of
the SMURF algorithm are given in Section~\ref{sec:algorithm}. The
paper is concluded in Section~\ref{sec:examples} with three detailed
test cases that span the majority of observation types likely to be
undertaken with \scuba, with an emphasis on the mitigation of
divergence problems, and characterizing the output maps: (i) Uranus, a
bright, compact source (Section~\ref{sec:point}); (ii) the Lockman
Hole, a blind survey of faint point-like source
(Section~\ref{sec:cosmo}); and (iii) the star-forming region M17,
including bright, extended emission (Section~\ref{sec:extended}). All
of the data analyzed in this paper are publicly available through the
CADC SCUBA-2 raw-data queries page for the dates and observation
numbers given in the
text\footnote{\texttt{http://www3.cadc-ccda.hia-iha.nrc-cnrc.gc.ca/
jcmt/search/scuba2}}. All of the analysis was performed using
\texttt{Starlink} version ... .

@@

Example data from the 450\,\micron\ subarray s4a where pickup appears
to be significant (observation 16 on 20100228) are shown in
Fig.~\ref{fig:magpickup}. The time-series for two bolometers in the
same column (not flatfielded) show that there is a strong signal with
a similar shape, but opposite signs. This behaviour is seen across the
entire array. The dark squid signal for the same columns exhibits a
similar shape and amplitude. Since the dark squid has no thermal
absorber or TES attached to it, this observed signal is not likely to
be optical or thermal in nature (although there can be some cross-talk
with the bolometers). Due to the fact that the sign of the gain in
each stage of SQUID amplification is random (although the combined
gain is constrained to be negative), and since magnetic field pickup
is only seen at the input to the second stage, the pickup can appear
with random signs for bolometers along a a column, giving it a
distinct signature from other common signals that appear with the same
sign.

@@

Significant experimentation with different models for noise sources
during \scuba\ commissioning required a highly flexible software
framework, and a configurable user interface. To achieve this goal,
while minimizing development time, it was decided to build SMURF as a
Starlink package \citep{2009ASPC..411..418J}, which provides access to
a large suite of libraries (including a commanding and messaging
interface, astrometric coordinate conversions and generation of
standard WCS information, file formats etc.). Furthermore, Starlink is
already used extensively at the Joint Astronomy Centre (host of the
JCMT) for many other systems, which helps with
interoperability. Though originally written in FORTRAN, many of the
core Starlink libraries are now ported to native C, or at least have a
C interface. It was therefore decided to develop SMURF in C as well
(rather than C++, for example, which would have required adding
further dependencies to Starlink), although we have taken an
object-oriented approach. For example, all of the data for a given
noise model is encapsulated in a C structure, and there is a standard
interface for all functions that handle models (member data and
functions for the class, respectively). In this way it is easy to
extend SMURF with new models. Parallelisation is incorporated in the
most time-consuming low-level routines using threads (e.g. performing
Fast Fourier Transforms, re-gridding the data), usually handling
either independent blocks of bolometers, or blocks of time, in each
thread, depending on the nature of the
calculation. \textbf{... something about scalability ...}

@@

First, the raw data are passed through a pre-processing stage which
corrects some of the more significant glitches, applies flat-field
corrections etc. Next, the iterative process begins (dashed box). Most
of the low-frequency noise is removed using common-mode
subtraction. Next, the extinction correction is applied (it is done at
this stage rather than during pre-processing to avoid multiplying
large atmospheric signals by any small errors in the flatfield
correction). At this stage the data resemble the grey bolometer traces
in Fig.~\ref{fig:bolos_mix}, with some residual baseline drifts still
visible. These drifts are removed using a high-pass filter implemented
with Fast Fourier Transforms (FFTs). The residual signal is then
re-gridded to estimate the map. Finally, the map is used to estimate
and remove the astronomical signal contamination in the bolometer
data, leaving a relatively clean data set in which to measure the
white noise level of each bolometer. Since the common-mode and
filtering stages will have introduced ringing in the map in the
vicinity of bright sources, the entire process is iterated. Each
component in the dashed box of Fig.~\ref{fig:dimm} is re-calculated in
sequence. In this way, the second time the common-mode is calculated,
for example, most of the bright astronomical sources in the data have
already been removed in the previous iteration when the map was
estimated and subtracted, reducing the amount of ringing in the map.

In Sections~\ref{sec:dataprep} and \ref{sec:components} we describe
the data pre-processing steps and iterative algorithm in
detail. Section~\ref{sec:converge} explores convergence tests and
degeneracies between model components.

@@

Prior to executing the iterative part of the SMURF algorithm, the data
must undergo several pre-processing steps. First, the data files are
read into memory and concatenated into continuous arrays (\scuba\ data
are broken up and written to disk every $\sim$30\,s). As the data are
loaded, the flatfield correction is also applied \citep[see
Section~2.1 in][]{dempsey2012}. Next, a series of data cleaning and
filtering procedures are applied; these include the removal of large
glitches that may hinder the iterative solution from converging, or
simply tasks that do not need to be iterated.

@@

The method used by SMURF to down-sample is to average together
multiple samples from the original time-series, $x_i$ to estimate the
lower-frequency output time-series, $y_j$. In general there will not
be an integer number of samples from $x_i$ in each of the $y_j$, so
fractional samples from $x_i$ are used at the $y_j$ time
boundaries. The algorithm is fast, since each sample in $x_i$ need
only be visited once. From a spectral point of view, this boxcar
average is equivalent to applying a sinc-function low-pass filter in
the frequency domain. Such behaviour is desirable since it serves as
an anti-aliasing filter; most non-white features above the Nyquist
frequency in $y_j$ that could contaminate the output data are largely
removed.

@@

\item As an alternative to the iterative map-making procedure,
cleaning by principal component analysis (PCA) is available as an
experimental pre-processing option. As described in
Section~\ref{sec:pca}, a new set of basic vectors for the bolometer
data are found which diagonalize their covariance matrix. The
brightest components can then be subtracted from the data. A single
parameter specifies the threshold on the amplitude of the eigenvalues
to be removed, as a number of standard deviations away from the mean
value. Subarrays are cleaned independently once for the full length of
each continuousy chunk of data. Given the computational expense of
this method, and initial tests which showed little improvement over
simple high-pass filtering, this method has not yet been explored in
detail with \scuba\ data. However, since systematic effects seem to
come and go, it is possible that PCA could be useful for particular
data sets.

@@

Sudden steps can occur in the time stream data from each bolometer,
with the most likely cause being cosmic ray events \citep[see
Section~3.5.3 in][]{holland2012}. The black curves in
Fig.~\ref{fig:steps1} show examples of such steps in the time streams
for two bolometers. If not removed, these steps can cause severe
ringing in the FLT model (Section~\ref{sec:flt}) and visible streaks
in the final map, corresponding to the paths of individual bolometers
over the sky.

Steps occur with a wide range of heights and shapes. The ratio of step
height to noise can vary from less than 10 to several hundred. Some
steps occur over a single sample, such as the step close to sample
5000 in the right-hand plot of Fig.~\ref{fig:steps1}, but others
happen more gradually, such as the step close to sample 5300. In
addition, a step can be preceded or followed by a short period of
instability, as is visible at the bottom of the step in the left hand
plot of Fig.~\ref{fig:steps1}. Further problems are caused by steps
that occur close together in time, such as the large downward step
followed by a smaller upward step close to sample 5000 in the right
hand plot of Fig.~\ref{fig:steps1}.

@@

We express the signal observed by the $i$th bolometer as a function of
time,
%
\begin{equation}
\mathbf{b}_i(t) = g_i[\mathbf{e}_i(t) \mathbf{a}_i(t) + \mathbf{n}_i(t)],
\end{equation}
%
where $\mathbf{a}(t)$ is the time-varying signal produced by scanning
the telescope across astronomical sources, $\mathbf{e}(t)$ is the
time-varying extinction, which is a function of the telescope
elevation and atmospheric conditions, and $\mathbf{n}_i(t)$ represents
sources of noise. The two terms in square brackets, as written, have
units of power delivered to the detectors (pW). The scale factor $g_i$
converts this effective power to the digitized units recorded on disk
(DAC) -- the \emph{flatfield} multiplied by a digitization constant --
and in this formulation it is assumed to be constant in time.

We then express the noise, $\mathbf{n}_i(t)$, as the sum of several
components,
%
\begin{equation}
  \mathbf{n}_i(t) = \mathbf{n}^\mathrm{w}_i(t) + \mathbf{n}^\mathrm{f}_i(t) +
  \mathbf{n}^\mathrm{c}(t),
\label{eq:noise}
\end{equation}
%
where $\mathbf{n}^\mathrm{w}_i(t)$ is uncorrelated white (thermal)
noise, $\mathbf{n}^\mathrm{f}_i(t)$ is low-frequency noise that is
\emph{not} correlated from bolometer-to-bolometer, and
$\mathbf{n}^\mathrm{c}(t)$ is a (predominantly low-frequency)
correlated or \emph{common-mode} component. The primary goal of
map-making is to model and remove $\mathbf{n}^\mathrm{f}_i$ and
$\mathbf{n}^\mathrm{c}$ from the time-series, and then by re-gridding
the remaining data we can hope to approach the theoretical noise limit
in the map. This procedure is summarized in Fig.~\ref{fig:dimm}.

 and run through a pre-processing stage that performs
several operations including spike and sudden level step removal to
enforce continuity in the time-series. Then, the iterative process
within the dashed box begins. It is generally assumed that the
dominant noise component is the common-mode, which is presumably
dominated by large variations in the fridge temperature and sky noise
(as argued in Section~\ref{sec:pca}). This signal is estimated as the
average signal observed by all of the bolometers at each time step,
and then removed from all of the bolometers. Once this large signal
has been removed, we apply the extinction correction, which is
determined completely from external measurements
\citep[see][]{dempsey2012}. Next, residual noise that differs from
bolometer-to-bolometer (primarily low-frequency) is removed, usually
using a high-pass filter. The resulting ``clean'' time-series are then
re-gridded to produce a map estimate using nearest-neighbour
sampling. Since each map pixel is calculated as the average of many
bolometer samples, it is fundamentally much less noisy than the raw
bolometer time-series. We can therefore project the map back into the
time domain (i.e. scanning the array across the current estimate of
the map), and create approximately noiseless estimates of what each
bolometer would have seen in the absence of noise sources, and remove
that from the time-series. This final step in the iteration gives a
residual signal (with low-frequency and astronomical noise components
removed) that may be used to measured the white noise level in each
detector, and also to estimate an approximate $\chi^2$ for the model
to track convergence. As we will discuss in the next sections, each
signal component estimate is biased slightly due to the presence of
signal from other components, so the entire process is
iterated. However, before the next iteration begins, we first undo the
multiplicative factors introduced by the extinction correction, and
also relative gains that may optionally be determined in the
common-mode estimation phase.

In practice, the map-maker is highly configurable, both in terms of
the order in which model components are fit (and whether they are fit
at all), and the details of \emph{how} they are fit. A summary of the
model components is given in the next section.

In addition to specifying when each model component is computed
(normally an attempt is made to order them by decreasing amplitude in
the time-series), there are a number of parameters that control their
operation. Table~\ref{tab:components} lists the model components and
the typical order in which they are calculated. The following sections
describe their implementation in detail.

@@

The \model{COM} model is the common-mode signal itself. It is a single
time series that is estimated by finding the mean of all bolometers at
each time step. Bolometer values that have been flagged as unusable
are excluded from the mean. {\bf Per: has a comment in
here about stability of the gain pattern... it should be constant
right?}.

Even after flat-fielding, bolometers may have slightly varying
sensitivities and so the amplitude of the common-mode variations will
also vary from bolometer to bolometer. Comparing each bolometer time
stream with the common-mode signal allows an estimate of the relative
bolometer sensitivity to be obtained. In practice, a least squares linear
fit is performed between the bolometer time series and the common-mode to
determine a gain and offset for each bolometer.  The gain and offset for
each bolometer is known as the \model{GAI} model. Each bolometer value
can then optionally be scaled and shifted using these values so that all
bolometers share a common (but as yet unknown) calibration. This provides
an alternative (or additional) flat-fielding strategy to that described
in section~\ref{sec:bolos}.

An option exists to cater for time-varying sensitivities. If used, the
least squares fits described above are performed on short blocks of
contiguous time slices, providing multiple gain and offset values for
each bolometer (one pair for each block of time slices). The gain and
offset at any required time slice can then be found by interpolation
between these values.

It can be seen from Fig.~\ref{fig:com} that some bolometers depart
radically from the common-mode, indicating some problem within the
bolometer. Such bolometers are identified by calculating the Pearson
coefficient of correlation between each bolometer time-stream and the
common mode. Bolometers for which the correlation coefficient is below
a specified limit, or which have unusually high or low gain compared to
the other bolometers, are flagged as bad in order to omit them from the
final map. If the option described
above for handling time-varying sensitivities is used, then a
correlation coefficient can be determined for each individual block of
time slices.  This allows individual bad blocks to be rejected from a
bolometer time stream, rather than rejecting the whole bolometer. This
is beneficial for bolometers that follow the common-mode for most of
the time but suffer from transient problems at specific times. {\bf
Per: If magnetic pickup dominates the common mode a negative gain is
normal and should not be removed. This might be a risk at high
elevation in very good stable weather!}

>>>>

\author[Edward~L.~Chapin~et~al.]{
  \parbox[t]{\textwidth}{
    Edward~L.~Chapin$^{1,2}$\thanks{E-mail:~echapin@sciops.esa.int,
      Present address: XMM SOC, ESAC, Apartado 78, 28691 Villanueva de
      la Ca\~nada, Madrid, Spain},
    David~S.~Berry$^{2}$,
    Andrew~G.~Gibb$^{1}$,
    Tim~Jenness$^{2}$,
    Douglas~Scott$^{1}$,
    Frossie~Economou$^2$\thanks{Present address: National Optical
      Astronomy Observatory, 950 N.\ Cherry Avenue, Tucson, AZ 85719, USA},
    Remo~P.~J.~Tilanus$^{2,3}$
  }
  \\
  \\
  $^{1}$Dept. of Physics \& Astronomy, University of British Columbia,
  6224 Agricultural Road, Vancouver, B.C. V6T 1Z1, Canada\\
  $^{2}$JointAstronomy Centre, 660 N. A`oh\={o}k\={u} Place, University
  Park, Hilo, Hawaii 96720, USA\\
  $^{3}$Netherlands Organisation for Scientific Research,
  Laan van Nieuw Oost-Indie 300, NL-2509 AC The Hague, The Netherlands}
\begin{document}

@@

\begin{abstract}
  The Submillimetre Common User Bolometer Array 2 (\scuba) is an
  instrument operating on the 15-m James Clerk Maxwell Telescope,
  consisting of about 5000 bolometers in each of two simultaneous
  imaging bands centered over 450 and 850\,\micron. The camera is
  scanned across the sky and records data at a rate of 200\,Hz. The
  data rate is much greater than that of previously existing
  submillimetre cameras, and represents a significant data anlaysis
  challenge. We describe the production of \scuba\ maps using the
  Sub-Millimetre User Reduction Facility (SMURF) in which we have
  adopted a fast, iterative approach to map-making that enables data
  reduction on single, modern, high-end desktop computers with
  execution times that are shorter than the observing times.  SMURF is
  used in an automated setting both at the telescope for real-time
  feedback to observers, as well as for the production of generic
  science products at the Canadian Astronomical Data Centre. A suite
  of tools are also provided for interactive data analysis.  Three
  detailed case studies are used to: (i) explore convergence
  properties of the map-maker using simple prior constraints (Uranus
  -- a point source); (ii) achieve the white-noise limit of the
  instrument for faint point-source studies (blank-field survey of the
  Lockman Hole); and (iii) demonstrate that our strategy is capable of
  recovering angular scales up to the size of the array footprint
  (approximately 5\,arcmin) for bright extended sources (star-forming
  region M17).
\end{abstract}

@@

The best existing map-making strategies for recovering information on
all angular scales are \emph{maximum likelihood} techniques, in which
the time-series data are expressed as a sampling of the ``true'' map
of the sky plus noise, and then the equation is inverted to estimate
the map as some weighted linear combination of the data that minimizes
the variance.  The first good description of this method appears in
\citet{janssen1992} for the production of maps from the COsmic
Background Explorer (COBE -- the description is relevant despite the
fact that it used a differential radiometer instead of
bolometers). Other descriptions in the CMB literature include
\citet{tegmark1997} and \citet{stompor2002}, and an application to
data from SCUBA is described in \citet{borys2004}.  The downside to
such methods is that they can be both computationally expensive and
have large memory requirements. While for some experimental designs
fast iterative methods for the inversion do exist with minimal memory
usage \citep[e.g.][]{wright1996}, for the more general map-making
problem involving many detectors, the problem is significantly
complicated both by the need to measure the cross power-spectra for
all unique pairs of bolometers, and then to perform the inversion
itself. The most promising maximum-likelihood method that may one day
be applied to \scuba\ data is ``SANEPIC'', which successfully reduced
maps from the Balloon-borne Large-Aperture Submillimeter Telescope,
involving hundreds of bolometers, and correctly incorporating
inter-bolometer noise correlations \citep[][]{patanchon2008}.

@@

This paper is divided in the following way. We first describe the
properties of \scuba\ data and scan patterns in Section~\ref{sec:data}
to motivate the data model that we have adopted. Note that most of the
analysis described here applies equally to all of the data taken
to-date with \scuba, including during the Shared-Risks Observing
period (S2SRO) from 2010 February 11 to 2010 March 25 (during which
only single subarrays were available at each of the two observing
wavelengths). A comparison between the S2SRO and fully-commissioned
instrument is given in Section~\ref{sec:s2sro}. Next, the details of
the SMURF algorithm are given in Section~\ref{sec:algorithm}. The
paper is concluded in Section~\ref{sec:examples} with three detailed
test cases that span the majority of observation types likely to be
undertaken with \scuba, with an emphasis on the mitigation of
divergence problems, and characterizing the output maps: (i) Uranus, a
bright, compact source (Section~\ref{sec:point}); (ii) the Lockman
Hole, a blind survey of faint point-like source
(Section~\ref{sec:cosmo}); and (iii) the star-forming region M17,
including bright, extended emission (Section~\ref{sec:extended}). All
of the data analyzed in this paper are publicly available through the
CADC \scuba\ raw-data queries page for the dates and observation
numbers given in the
text\footnote{\texttt{http://www3.cadc-ccda.hia-iha.nrc-cnrc.gc.ca/
jcmt/search/scuba2}}. All of the analysis was performed using
\texttt{Starlink} version ... .

@@

Example data from the 450\,\micron\ subarray s4a where pickup appears
to be significant (observation 16 on 20100228) are shown in
Fig.~\ref{fig:magpickup}. The time-series for two bolometers in the
same column (not flatfielded) show that there is a strong signal with
a similar shape, but opposite signs. This behaviour is seen across the
entire array. The dark squid signal for the same columns exhibits a
similar shape and amplitude. Since the dark squid has no thermal
absorber or TES attached to it, this observed signal is not likely to
be optical or thermal in nature (although there can be some cross-talk
with the bolometers). Due to the fact that the sign of the gain in
each stage of SQUID amplification is random (although the combined
gain is constrained to be negative), and since magnetic field pickup
is only seen at the input to the second stage, the pickup can appear
with random signs for bolometers along a column, giving it a distinct
signature from other common signals that appear with the same sign.

@@

Significant experimentation with different models for noise sources
during \scuba\ commissioning required a highly flexible software
framework, and a configurable user interface. To achieve this goal,
while minimizing development time, it was decided to build SMURF as a
Starlink package \citep{2009ASPC..411..418J}, which provides access to
a large suite of libraries (including a commanding and messaging
interface, astrometric coordinate conversions and generation of
standard WCS information, file formats etc.). Furthermore, Starlink is
open source, and already used extensively at the Joint Astronomy
Centre (host of the JCMT) for many other systems, which helps with
interoperability. Though originally written in FORTRAN, many of the
core Starlink libraries are now ported to native C, or at least have a
C interface. It was therefore decided to develop SMURF in C as well
(rather than C++, for example, which would have required adding
further dependencies to Starlink), although we have taken an
object-oriented approach. For example, all of the data for a given
noise model is encapsulated in a C structure, and there is a standard
interface for all functions that handle models (member data and
functions for the class, respectively). In this way it is easy to
extend SMURF with new models. Parallelisation is incorporated in the
most time-consuming low-level routines using threads (e.g. performing
Fast Fourier Transforms, re-gridding the data), usually handling
either independent blocks of bolometers, or blocks of time, in each
thread, depending on the nature of the
calculation. \textbf{... something about scalability ...}

@@

First, the raw data are passed through a pre-processing stage which
corrects some of the more significant glitches, applies flat-field
corrections etc. Next, the iterative process begins (dashed box). Most
of the low-frequency noise is removed using common-mode
subtraction. Next, the extinction correction is applied (it is done at
this stage rather than during pre-processing to avoid multiplying
large atmospheric signals by any small errors in the flatfield
correction). At this stage the data resemble the grey bolometer traces
in Fig.~\ref{fig:bolos_mix}, with some residual baseline drifts still
visible. These drifts are removed using a high-pass filter implemented
with Fast Fourier Transforms (FFTs). The residual signal is then
re-gridded to estimate the map. Finally, the map is used to estimate
and remove the astronomical signal contamination in the bolometer
data, leaving a relatively clean data set in which to measure the
white noise level of each bolometer. Since the common-mode and
filtering stages will have introduced ringing in the map in the
vicinity of bright sources, the entire process is iterated. Each
component in the dashed box of Fig.~\ref{fig:dimm} is re-calculated in
sequence. In this way, the second time the common-mode is calculated,
for example, most of the bright astronomical sources in the data have
already been removed in the previous iteration when the map was
estimated and subtracted, reducing the amount of ringing in the map
once it is re-estimated. Note that the extinction and gain denoted in
a second dashed box are \emph{multiplicative} rather than additive
models, and so they must be inverted prior to the start of each
iteration to preserve the units (see Sections~\ref{sec:comgai} and
\ref{sec:ext}).

In general, SMURF is highly configurable, including many options for
both the pre-processing stage and the iterative model components
(which models are used, what order they are calculated in, how they
are calculated).  In Sections~\ref{sec:dataprep} and
\ref{sec:components} we describe the data pre-processing steps and
iterative algorithm in detail. Section~\ref{sec:converge} explores
convergence tests and degeneracies between model components.

@@

Prior to executing the iterative part of the algorithm, the data must
undergo several pre-processing steps. First, the data files are read
into memory and concatenated into continuous arrays (\scuba\ data are
broken up and written to disk every $\sim$30\,s during data
acquisition). As the data are loaded, they are multiplied by the
flatfield correction \citep[see Section~2.1 in][]{dempsey2012}. Next,
a series of data cleaning and filtering procedures are applied; these
include the removal of large glitches that may hinder the iterative
solution from converging, or simply tasks that do not need to be
iterated.

@@

The method used by SMURF to down-sample is to average together
multiple samples from the original time-series, $x_i$ to estimate the
lower-frequency output time-series, $y_j$. In general there will not
be an integer number of samples from $x_i$ in each of the $y_j$, so
fractional samples from $x_i$ are used at the $y_j$ time
boundaries. The algorithm is fast, since each sample in $x_i$ need
only be visited once. From a spectral point of view, this boxcar
average is equivalent to applying a sinc-function low-pass
filter. Such behaviour is desirable since it serves as an
anti-aliasing filter; most non-white features above the Nyquist
frequency in $y_j$ that could contaminate the output data are largely
removed.

@@

\item As an alternative to the iterative map-making procedure,
cleaning by principal component analysis (PCA) is available as an
experimental pre-processing option (Section~\ref{sec:pca}). The most
significant components in the analysis are identified, and the
projection of each bolometer time series along their eigenvectors are
removed. A single parameter specifies the threshold on the amplitude
of the eigenvalues to be removed, as a number of standard deviations
away from the mean value. Subarrays are cleaned independently, once,
for the full length of each continuousy chunk of data. Given the
computational expense of this method, and initial tests which showed
little improvement over simple high-pass filtering, this method has
not yet been explored in detail with \scuba\ data. However, since
systematic effects seem to come and go, it is possible that PCA could
be useful for particular data sets.

@@

Sudden steps can occur in the time stream data from each bolometer,
with the most likely cause being cosmic ray events \citep[see
Section~3.5.3 in][]{holland2012}. The black curves in
Fig.~\ref{fig:steps1} show examples of such steps in the time streams
for two bolometers. If not removed, these steps can cause severe
ringing when filtering, and visible streaks in the final map,
corresponding to the paths of individual bolometers over the sky.

Steps occur with a wide range of heights and shapes. The ratio of step
height to noise can vary from less than 10 to several hundred. Some
steps occur over a single sample, such as the step close to sample
5000 in the right-hand plot of Fig.~\ref{fig:steps1}, but others
happen more gradually, such as the step close to sample 5300. In
addition, a step can be preceded or followed by a short period of
instability, as is visible at the bottom of the step in the left hand
plot of Fig.~\ref{fig:steps1} (this is probably due to the response of
the \scuba\ anti-aliasing filter; the sudden large step occurs prior
to the 200\,Hz re-sampling). Further problems are caused by steps that
occur close together in time, such as the large downward step followed
by a smaller upward step close to sample 5000 in the right hand plot
of Fig.~\ref{fig:steps1}.

@@

Once the data have been cleaned, and the worst data flagged to be
ignored, the iterative solution for the map and contaminating noise
signals begins.

First, we describe the generic model for \scuba\ data. We express the
signal observed by the $i$th bolometer as a function of time,
%
\begin{equation}
\mathbf{b}_i(t) = f_i[\mathbf{e}_i(t) \mathbf{a}_i(t) + \mathbf{n}_i(t)],
\label{eq:model}
\end{equation}
%
where $\mathbf{a}(t)$ is the time-varying signal produced by scanning
the telescope across astronomical sources, $\mathbf{e}(t)$ is the
time-varying extinction, which is a function of the telescope
elevation and atmospheric conditions, and $\mathbf{n}_i(t)$ represents
sources of noise. The two terms in square brackets, as written, have
units of power delivered to the detectors (pW). The scale factor $f_i$
converts this effective power to the digitized units recorded on disk
(DAC) -- the flatfield multiplied by a digitization constant (applied
once during pre-processing) -- which in this formulation is assumed to
be constant in time.

We then express the noise, $\mathbf{n}_i(t)$, as the sum of several
components,
%
\begin{equation}
  \mathbf{n}_i(t) = \mathbf{n}^\mathrm{w}_i(t) +
  g_i\mathbf{n}^\mathrm{c}(t) + \mathbf{n}^\mathrm{f}_i(t),
\label{eq:noise}
\end{equation}
%
where $\mathbf{n}^\mathrm{w}_i(t)$ is uncorrelated white noise,
$\mathbf{n}^\mathrm{c}(t)$ is a correlated or \emph{common-mode}
signal (with an optional scale factor $g_i$ for each bolometer), and
$\mathbf{n}^\mathrm{f}_i(t)$ is (predominantly low-frequency) noise in
excess of the white noise level that is either uncorrelated from
bolometer-to-bolometer, or at least does not have a simple correlation
relationship that would lead to it being included in
$\mathbf{n}^\mathrm{c}(t)$. The primary goal of map-making is to model
and remove $\mathbf{n}^\mathrm{c}(t)$ and $\mathbf{n}^\mathrm{f}_i(t)$
from the time-series, and producing a map from what is left.  This
procedure is summarized in Fig.~\ref{fig:dimm}.

Each model component is estimated from the data in a user-defined
sequence, where the typically strongest components are estimated
first, and then subtracted, leaving a cleaner signal for subsequent
model estimates. One exception to this rule is the extinction
correction, $\mathbf{e}_i(t)$ in Eq.~\ref{eq:model}, which is simply a
multiplicative factor that is known prior to map-making. Note that the
order in which this correction occurs will modify the units of
subsequently calculated additive model components (e.g. additional
brackets may be needed in Eq.~\ref{eq:model} after
$\mathbf{e}_i(t)$). Another exception is the calculation of
$\mathbf{n}^\mathrm{c}(t)$ and $\mathbf{n}^\mathrm{f}_i(t)$ in which
the previous iterations are usually added back into the time-series
simultaneously to avoid degeneracies (see Section~\ref{sec:converge}).

Usually the last two components estimated are $\mathbf{a}_i(t)$ and
$\mathbf{w}_i(t)$ once the other contaminating noise sources, and
extinction, have been corrected for. Provided that these two signals
are the only components of the residual bolometer signal, regridding
into a map provides the highest \snr\ estimate of the astronomical
sky. The signal produced by this map is then projected back into the
time-domain to estimate and remove $\mathbf{a}_i(t)$, leaving a
residual signal that should be dominated by white noise,
$\mathbf{n}_i^\mathrm{w}(t)$.

After the first iteration, there will almost certainly be correlated
errors between the model estimates. For example, the presence of a
bright astronomical source will contaminate $\mathbf{n}^\mathrm{c}(t)$
(which is a simple average of all of the bolometer time-series at each
instant in time), leading to an \emph{over-subtraction}, and in turn,
negative bowls in the map.

Subsequent iterations, however, diminish such problems. Each additive
model component is re-estimated in the same sequence, after first
adding the previous estimate back into the time-series (but
importantly, not the other components). In this case, much of the
astronomical signal will have been identified and removed in the
previous calculation of $\mathbf{a}_i(t)$, and therefore there will be
less contamination in $\mathbf{n}^\mathrm{c}(t)$, and less negative
bowling in the map. Any multiplicative models (e.g.,
$\mathbf{e}_i(t)$) are inverted immediately prior to the start of a
new iteration to preserve the units of the data. The iterative
solution will thus converge, barring degeneracies between model
components.

The full set of model components available, and the parameters that
control them, are described in the following
sections. Table~\ref{tab:components} shows the typical order in which
they are calculated. For a discussion on convergence tests and
degenerices, see Section~\ref{sec:converge}.

@@

The \model{COM} model is the common-mode signal itself
($\mathbf{n}^{c}(t)$ in Eq.~\ref{eq:noise}). It is a single time
series that is estimated by finding the mean of all bolometers at each
time step. Bolometer values that have been flagged as unusable are
excluded from the mean.

Even after flat-fielding, bolometers may have slightly varying
sensitivities and so the amplitude of the common-mode variations will
also vary from bolometer to bolometer. Comparing each bolometer time
stream with the common-mode signal allows an estimate of the relative
bolometer sensitivity to be obtained. In practice, a least squares
linear fit is performed between the bolometer time series and the
common-mode to determine a gain ($g_i$ in Eq.~\ref{eq:noise}) and
offset for each bolometer.  The gain and offset for each bolometer is
known as the \model{GAI} model. Each bolometer value can then
optionally be scaled and shifted using these values so that all
bolometers share a common (but as yet unknown) calibration. This
provides an alternative (or additional) flat-fielding strategy to that
described in section~\ref{sec:bolos}.

An option exists to cater for time-varying sensitivities. In
principle, the gain of the bolometers should be constant in
time. However, there is evidence that there are slight variations, and
this option does tend to slightly improve the noise in maps.  If used,
the least squares fits described above are performed on short blocks
of contiguous time slices, providing multiple gain and offset values
for each bolometer (one pair for each block of time slices). The gain
and offset at any required time slice can then be found by
interpolation between these values.

It can be seen from Fig.~\ref{fig:com} that some bolometers depart
radically from the common-mode, indicating some problem within the
bolometer. Such bolometers are identified by calculating the Pearson
coefficient of correlation between each bolometer time-stream and the
common mode. Bolometers for which the correlation coefficient is below
a specified limit, or which have unusually high or low gain compared
to the other bolometers, are flagged as bad in order to omit them from
the final map. If the option described above for handling time-varying
sensitivities is used, then a correlation coefficient can be
determined for each individual block of time slices.  This allows
individual bad blocks to be rejected from a bolometer time stream,
rather than rejecting the whole bolometer. This is beneficial for
bolometers that follow the common-mode for most of the time but suffer
from transient problems at specific times. As a warning, rejecting
data based on outlier gain values can be misleading in cases where the
data are dominated by magnetic field pickup. For example, the
common-mode signal for the data in Fig.~\ref{fig:magpickup} would
resemble the azimuthal scan pattern, but clearly the gains will have
opposite signs (such that one or the other bolometer would be
erroneously flagged).
